project-plan

### 2025-09-25 project-plan.qmd
Addin comments into participants
Adding comments

Management of TBI in the ICU 
------------

Initial management of TBI spans prehospital and early in-hospital phases and centres on rapid physiological stabilisation and diagnostics in order for secondary injuries to be limited. Protection of the airway in patients with impaired consciousness, adequate oxygenation and normocapnia, and fast correction of hypotension are all prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased mortality (6–9).  Non-contrast head CT should be performed without delay to identify surgically remediable lesions. Depending on mechanism and physiology, whole-body CT may be used selectively, as randomised data show no mortality advantage over targeted imaging, while observational work underscores the benefit of immediate CT pathways for severely injured patients (10,11). Haemostatic strategies in the early stages, such as administering tranexamic acid within three hours of injury, can reduce head injury-related mortality (12). After this initial phase, depending on the patient's condition, management transitions to other areas of the hospital, where sustained prevention of secondary systemic and intracranial insults becomes the primary focus.

@@

(At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken. For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended. In cases where there are signs of impending herniation, the handbook advises measures such as controlled hyperventilation and osmotherapy with mannitol or hypertonic saline. All this is always in conjunction with urgent neurosurgical evaluation. (Källa Karolinska handboken))  <!-- Remove? -->

@@

Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors, deviations from established practice, and treatment delays, all of which may negatively influence outcomes. In response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted benchmarking system designed to assess the quality of trauma care. Through audit filters and peer review, it identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11). Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid management (12–16). Importantly, some of these findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain vulnerable to error. Furthermore, injuries to the central nervous system account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma Registry (16). Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.

@@

The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events such as falls, traffic incidents, or other external violence. Since 2024 , patients have been included if a trauma team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS) greater than fifteen without trauma team activation. Patients transferred to a hospital within seven days of the traumatic event with a NISS greater than fifteen are also included. Exclusion criteria are cases where the only traumatic condition is chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with protected identity, and patients declared dead on arrival. Since these inclusion and exclusion criteria have changed over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period (Källa SweTrau årsbok 2024).

@@

Participants
------------
Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and 2023 who also had been screened for opportunity for improvements. Furthermore, TBI was defined as (a GCS score of ≤8 combined with an Abbreviated Injury Scale ≥3 for the head region ….  ) Patients were eligible irrespective of level of care and were not restricted to intensive care admissions.

Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality review processes that differ from those for adults . Also excluded were patients declared dead on arrival, since assessment of opportunities for improvement in these cases is not relevant to the aim of this study.

@@

The primary outcome of this study was the presence of one or more opportunities for improvement during the management of patients with TBI. As mentioned, OFIs were adjudicated after a screening process with audit filters and a subsequent review in multidisciplinary morbidity-and-mortality conferences. Following the peer review, the result of each case discussion was recorded in the trauma care quality register. Moreover, the outcome was presented as a binary variable, “Yes - At least one OFI identified” and “No - No OFI identified”.

@@

Furthermore, the categorical variables include…

Statistical methods
----------------

*Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*

>>>>

Management of TBI 
------------

Initial management of TBI spans prehospital and early in-hospital phases and centres on rapid physiological
stabilisation and diagnostics in order for secondary injuries to be limited. Protection of the airway in patients
with impaired consciousness, adequate oxygenation and normocapnia, and fast correction of hypotension are all
prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased
mortality (6–9).  Non-contrast head CT should be performed without delay to identify surgically remediable lesions.
Depending on mechanism and physiology, whole-body CT may be used selectively, as randomised data show no mortality
advantage over targeted imaging, while observational work underscores the benefit of immediate CT pathways for
severely injured patients (10,11). Haemostatic strategies in the early stages, such as administering tranexamic acid
within three hours of injury, can reduce mortality (12). After this initial phase, depending on the patient's condition, management transitions to other areas of the hospital, where sustained prevention of secondary systemic
and intracranial insults becomes the primary focus.

@@

(At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT
imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended. In cases
where there are signs of impending herniation, the handbook advises measures such as controlled hyperventilation and
osmotherapy with mannitol or hypertonic saline. All this is always in conjunction with urgent neurosurgical
evaluation. (Källa Karolinska handboken))  <!-- Remove paragraph? -->

@@

Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors,
deviations from established practice, and treatment delays, all of which may negatively influence outcomes. In
response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted
benchmarking system designed to assess the quality of trauma care. Through audit filters and peer review, it
identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11). Previous
studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in
imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid
management (12–16). Importantly, some of these findings were observed in high-performing level I trauma centres,
underscoring that even mature systems remain vulnerable to error. Furthermore, injuries to the central nervous system
account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such
as the Arkansas Trauma Registry (16). Yet, while previous research has described OFI in trauma resuscitation and in
general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI
specifically among patients with TBI. <!-- Add a part after ACS TQIP using WHO as a reference to TQIP-->

@@

The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events
such as falls, traffic incidents, or other external violence. Since 2024 , patients have been included if a trauma
team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS)
greater than fifteen without trauma team activation. Patients transferred to a hospital within seven days of the
traumatic event with a NISS greater than fifteen are also included. Exclusion criteria are cases where the only
traumatic condition is chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with
protected identity, and patients declared dead on arrival. Since these inclusion and exclusion criteria have changed
over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period
(Källa SweTrau årsbok 2024). <!--Use the 2023 version instead? Data is up to 2023 -->

@@

Participants
------------
Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and
2023 who also had been screened for opportunity for improvements. Furthermore, TBI was defined as (a GCS score of ≤8
combined with an Abbreviated Injury Scale ≥3 for the head region ….  ) Patients were eligible irrespective of level <!-- Definition found in Elins paper, same applies?-->
of care and were not restricted to intensive care admissions.

Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality <!-- is this true? could not fins source-->
review processes that differ from those for adults . Also excluded were patients declared dead on arrival, since
assessment of opportunities for improvement in these cases is not relevant to the aim of this study.

@@

The primary outcome of this study was the presence of one or more opportunities for improvement during the management
of patients with TBI. As mentioned, OFIs were determined after a screening process with audit filters and a
subsequent review in multidisciplinary morbidity-and-mortality conferences. Following the peer review, the result of
each case discussion was recorded in the trauma care quality register and was presented as a binary variable, “Yes -
At least one OFI identified” and “No - No OFI identified”.

@@

Furthermore, the categorical variables include…
<!-- Choose variables, not sure which too chose-->

Statistical methods
----------------
<!-- unsure -->

<!-- *Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).* -->
